---
title: "Getting started with lbstproj"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with lbstproj}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
tmp_proj_dir <- fs::path(tempdir(), "fake-trial")
fs::dir_create(tmp_proj_dir)

knitr::opts_knit$set(root.dir = tmp_proj_dir)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 5,
  fig.height = 4,
  out.width = "70%"
)
```

## Introduction

This tutorial walks through a complete `lbstproj` workflow using a fake oncology
project. We use the `trial` dataset — a simulated randomized clinical trial of
200 patients assigned to **Drug A** or **Drug B** — to demonstrate how to:

1. Scaffold a standardized project
2. Register outputs in the **Table of Tables** (TOT)
3. Write and run data preparation, figure, and table scripts
4. Generate a self-contained report

---

## 1. Create a project

`create_project()` builds the full directory structure and adds a `DESCRIPTION`
file and a blank Table of Tables. We pass `force = TRUE` and `open = FALSE` to
skip interactive prompts.

```{r project-setup, message=FALSE}
library(lbstproj)

create_project(
  path = tmp_proj_dir,
  title = "Fake oncology trial",
  client = "John Doe",
  department = "Oncology",
  author = "Jane Doe",
  email = "jane.doe@university.com",
  open = FALSE,
  force = TRUE
)
```

The project now has the following structure:

```{r dir-tree}
fs::dir_tree(tmp_proj_dir, recurse = TRUE)
```

| Folder            | Purpose |
|-------------------|---------|
| `data/raw/`       | Raw input files (CSV, Excel, …) |
| `data/processed/` | Cleaned datasets saved as `.rds` |
| `data/tables/`    | GT tables saved as `.rds` |
| `data/tot/`       | The Table of Tables (Excel + cached `.rds`) |
| `R/data/`         | Data preparation scripts |
| `R/figures/`      | Figure scripts |
| `R/tables/`       | Table scripts |
| `results/figures/`| Final PNG/PDF figures |
| `results/tables/` | Word tables (`.docx`) |
| `report/`         | Quarto report |

---

## 2. Import raw data

The `trial` dataset is bundled with `lbstproj` as a CSV. Copy it to `data/raw/`:

```{r import-raw, echo=FALSE}
fs::file_copy(
  fs::path_package("lbstproj", "extdata", "trial.csv"),
  fs::path(tmp_proj_dir, "data", "raw", "trial.csv"),
  overwrite = TRUE
)
```

```r
# In a real project you would place your raw data file in data/raw/ manually.
fs::file_copy("path/to/trial.csv", "data/raw/trial.csv")
```

```{r show-raw-data}
trial_raw <- read.csv("data/raw/trial.csv")
head(trial_raw)
```

The dataset has `r nrow(trial_raw)` patients with eight variables: treatment arm
(`trt`), `age`, biomarker level (`marker`), tumour stage (`stage`), histological
grade (`grade`), binary `response`, and survival endpoints (`death`, `ttdeath`).

---

## 3. The Table of Tables

The **Table of Tables** (TOT) is the central registry of every output your
project will produce. It lives in `data/tot/table_of_tables.xlsx` and has four
columns:

| Column    | Description |
|-----------|-------------|
| `id`      | Short unique identifier (referenced inside scripts) |
| `type`    | `"figure"` or `"table"` |
| `name`    | File name used for saving (letters, numbers, hyphens only) |
| `caption` | Caption that appears in the report |

For this project we plan two figures and two tables:

```{r show-tot-data}
tot_data <- data.frame(
  id = c("fig-01", "fig-02", "tab-01", "tab-02"),
  type = c("figure", "figure", "table", "table"),
  name = c("fig-age-dist", "fig-km", "tab-baseline", "tab-response"),
  caption = c(
    "Age distribution by treatment arm.",
    "Kaplan-Meier curve of the survival over time per treatment arm.",
    "Baseline patient characteristics by treatment arm.",
    "Tumour response rates by treatment arm."
  ),
  stringsAsFactors = FALSE
)
tot_data
```

```{r write-tot, echo=FALSE}
writexl::write_xlsx(
  list(tot = tot_data),
  path = fs::path(tmp_proj_dir, "data", "tot", "table_of_tables.xlsx")
)
```

After filling in the spreadsheet, call `import_tot()` to validate it and cache
it as an `.rds` file:

```{r import-tot}
import_tot()
```

You can always inspect the TOT with `load_tot()`:

```{r load-tot}
load_tot()
```

---

## 4. Create analysis scripts

`create_from_tot()` reads the TOT and generates stub R scripts for each entry.
Always do a **dry run** first to preview what will be created:

```{r create-from-tot-dry}
create_from_tot(dry_run = TRUE)
```

Once satisfied, run without `dry_run` to create the stubs:

```{r create-from-tot-actual}
create_from_tot(dry_run = FALSE)
```

Each stub is pre-filled with a header, `get_info()` to retrieve TOT metadata,
library calls, and placeholder comments:

```{r show-stub, echo=FALSE, comment=""}
cat(
  readLines(fs::path(tmp_proj_dir, "R", "figures", "fig-age-dist.R")),
  sep = "\n"
)
```

You fill in the placeholder sections with your analysis. The next sections show
the completed scripts for our oncology project.

---

## 5. Data preparation

All cleaning and reshaping happens in `R/data/`. The script below loads the raw
CSV, coerces categorical variables to factors, and saves the result with
`save_data()`.

```{r data-script-display, eval=FALSE}
# R/data/trial-clean.R

library(dplyr)

# Load raw data ---------------------------------------------------------------
raw <- read.csv("data/raw/trial.csv")

# Clean -----------------------------------------------------------------------
trial <- raw |>
  mutate(
    trt = factor(trt),
    stage = factor(stage),
    grade = factor(grade, levels = c("I", "II", "III"))
  )

# Save ------------------------------------------------------------------------
save_data(trial, name = "trial-clean")
```

```{r run-data-script, echo=FALSE, message=FALSE}
writeLines(
  c(
    'library(dplyr)',
    'raw <- read.csv("data/raw/trial.csv")',
    'trial <- raw |>',
    '  mutate(',
    '    trt   = factor(trt),',
    '    stage = factor(stage),',
    '    grade = factor(grade, levels = c("I", "II", "III"))',
    '  )',
    'lbstproj::save_data(trial, name = "trial-clean")'
  ),
  fs::path(tmp_proj_dir, "R", "data", "trial-clean.R")
)

withr::with_options(list(lbstproj.quiet = TRUE), {
  source(fs::path(tmp_proj_dir, "R", "data", "trial-clean.R"))
})
```

After running the script you can inspect the processed dataset:

```{r show-processed-data}
trial <- readRDS("data/processed/trial-clean.rds")
dplyr::glimpse(trial)
```

---

## 6. Figures

### Figure 1 — Age distribution

```{r fig-age-script-display, eval=FALSE}
# R/figures/fig-age-dist.R

library(dplyr)
library(ggplot2)

info <- get_info("fig-01")
trial <- readRDS("data/processed/trial-clean.rds")

fig <- ggplot(trial, aes(x = trt, y = age, fill = trt)) +
  geom_violin(alpha = 0.4, trim = FALSE) +
  geom_boxplot(width = 0.15, alpha = 0.8, outlier.shape = NA) +
  scale_fill_brewer(palette = "Set2") +
  labs(x = "Treatment arm", y = "Age (years)") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")

save_figure(fig, name = "fig-age-dist", width = 5, height = 4)
```

```{r run-fig-age, echo=FALSE, message=FALSE, warning=FALSE}
fig_age_script <- c(
  'library(dplyr)',
  'library(ggplot2)',
  'info  <- lbstproj::get_info("fig-01")',
  'trial <- readRDS("data/processed/trial-clean.rds")',
  'fig <- ggplot(trial, aes(x = trt, y = age, fill = trt)) +',
  '  geom_violin(alpha = 0.4, trim = FALSE) +',
  '  geom_boxplot(width = 0.15, alpha = 0.8, outlier.shape = NA) +',
  '  scale_fill_brewer(palette = "Set2") +',
  '  labs(x = "Treatment arm", y = "Age (years)") +',
  '  theme_minimal(base_size = 12) +',
  '  theme(legend.position = "none")',
  'lbstproj::save_figure(fig, name = "fig-age-dist", width = 5, height = 4)'
)
writeLines(
  fig_age_script,
  fs::path(tmp_proj_dir, "R", "figures", "fig-age-dist.R")
)

withr::with_options(list(lbstproj.quiet = TRUE), {
  source(fs::path(tmp_proj_dir, "R", "figures", "fig-age-dist.R"))
})
fig_age_dist <- fig
```

```{r show-fig-age, echo=FALSE, fig.cap="Age distribution by treatment arm."}
fig_age_dist
```

### Figure 2 — Kaplan-Meier curve

```{r fig-km-script-display, eval=FALSE}
# R/figures/fig-km.R

library(dplyr)
library(ggplot2)
library(ggsurvfit)

info <- get_info("fig-02")
trial <- readRDS("data/processed/trial-clean.rds")

mod <- survfit2(
  data = data,
  formula = Surv(ttdeath, death) ~ response
)

fig <- ggsurvfit(mod) +
  add_confidence_interval() +
  labs(
    x = "Months since treatment",
    y = "Survival probability",
    color = "Tumor Response",
    fill = "Tumor Response"
  ) +
  theme_minimal(base_size = 12)

save_figure(fig, name = "fig-km", width = 5, height = 4)
```

```{r run-fig-response, echo=FALSE, message=FALSE, warning=FALSE}
fig_resp_script <- c(
  'library(dplyr)',
  'library(ggplot2)',
  'library(ggsurvfit)',
  'info  <- lbstproj::get_info("fig-02")',
  'trial <- readRDS("data/processed/trial-clean.rds")',
  'mod <- survfit2(',
  '  data = trial,',
  '  formula = Surv(ttdeath, death) ~ response',
  ')',
  'fig <- ggsurvfit(mod) +',
  'add_confidence_interval() +',
  'labs(',
  '  x = "Months since treatment",',
  '  y = "Survival probability",',
  '  color = "Tumor Response",',
  '  fill = "Tumor Response"',
  ') +',
  'theme_minimal(base_size = 12)',
  'lbstproj::save_figure(fig, name = "fig-km", width = 5, height = 4)'
)
writeLines(
  fig_resp_script,
  fs::path(tmp_proj_dir, "R", "figures", "fig-km.R")
)

withr::with_options(list(lbstproj.quiet = TRUE), {
  source(fs::path(tmp_proj_dir, "R", "figures", "fig-km.R"))
})
fig_response <- fig
```

```{r show-fig-response, echo=FALSE, fig.cap="Tumour response rate by treatment arm."}
fig_response
```

---

## 7. Tables

Tables are saved as `gt` objects and optionally exported to Word. The workflow
below uses `gtsummary` to produce publication-ready summaries, then converts to
`gt` with `as_gt()` before saving.

### Table 1 — Baseline characteristics

```{r tab-baseline-script-display, eval=FALSE}
# R/tables/tab-baseline.R

library(dplyr)
library(gtsummary)

info <- get_info("tab-01")
trial <- readRDS("data/processed/trial-clean.rds")

tab <- trial |>
  select(trt, age, stage, grade) |>
  tbl_summary(
    by = trt,
    label = list(
      age ~ "Age (years)",
      stage ~ "Tumour stage",
      grade ~ "Grade"
    )
  ) |>
  add_overall() |>
  add_p() |>
  modify_header(label ~ "**Characteristic**") |>
  bold_labels() |>
  as_gt() |>
  gt::tab_header(title = info$caption)

save_table(tab, name = "tab-baseline", export = FALSE)
```

```{r run-tab-baseline, echo=FALSE, message=FALSE, warning=FALSE}
tab_bl_script <- c(
  'library(dplyr)',
  'library(gtsummary)',
  'info  <- lbstproj::get_info("tab-01")',
  'trial <- readRDS("data/processed/trial-clean.rds")',
  'tab <- trial |>',
  '  select(trt, age, stage, grade) |>',
  '  tbl_summary(',
  '    by    = trt,',
  '    label = list(',
  '      age   ~ "Age (years)",',
  '      stage ~ "Tumour stage",',
  '      grade ~ "Grade"',
  '    )',
  '  ) |>',
  '  add_overall() |>',
  '  add_p() |>',
  '  modify_header(label ~ "**Characteristic**") |>',
  '  bold_labels() |>',
  '  as_gt() |>',
  '  gt::tab_header(title = info$caption)',
  'lbstproj::save_table(tab, name = "tab-baseline", export = FALSE)'
)
writeLines(
  tab_bl_script,
  fs::path(tmp_proj_dir, "R", "tables", "tab-baseline.R")
)

withr::with_options(list(lbstproj.quiet = TRUE), {
  source(fs::path(tmp_proj_dir, "R", "tables", "tab-baseline.R"))
})
```

```{r show-tab-baseline}
readRDS("data/tables/tab-baseline.rds")
```

### Table 2 — Response rates

```{r tab-response-script-display, eval=FALSE}
# R/tables/tab-response.R

library(dplyr)
library(gtsummary)

info <- get_info("tab-02")
trial <- readRDS("data/processed/trial-clean.rds")

tab <- trial |>
  select(trt, response) |>
  mutate(
    response = factor(response, levels = c(1, 0), labels = c("Yes", "No"))
  ) |>
  tbl_summary(
    by = trt,
    label = list(response ~ "Tumour response")
  ) |>
  add_p() |>
  modify_header(label ~ "**Variable**") |>
  bold_labels() |>
  as_gt() |>
  gt::tab_header(title = info$caption)

save_table(tab, name = "tab-response", export = FALSE)
```

```{r run-tab-response, echo=FALSE, message=FALSE, warning=FALSE}
tab_resp_script <- c(
  'library(dplyr)',
  'library(gtsummary)',
  'info  <- lbstproj::get_info("tab-02")',
  'trial <- readRDS("data/processed/trial-clean.rds")',
  'tab <- trial |>',
  '  select(trt, response) |>',
  '  mutate(response = factor(response, levels = c(1, 0), labels = c("Yes", "No"))) |>',
  '  tbl_summary(',
  '    by    = trt,',
  '    label = list(response ~ "Tumour response")',
  '  ) |>',
  '  add_p() |>',
  '  modify_header(label ~ "**Variable**") |>',
  '  bold_labels() |>',
  '  as_gt() |>',
  '  gt::tab_header(title = info$caption)',
  'lbstproj::save_table(tab, name = "tab-response", export = FALSE)'
)
writeLines(
  tab_resp_script,
  fs::path(tmp_proj_dir, "R", "tables", "tab-response.R")
)

withr::with_options(list(lbstproj.quiet = TRUE), {
  source(fs::path(tmp_proj_dir, "R", "tables", "tab-response.R"))
})
```

```{r show-tab-response}
readRDS("data/tables/tab-response.rds")
```

---

## 8. Running all outputs

Once all scripts are written, `run_all_figures()` and `run_all_tables()` source
every script in `R/figures/` and `R/tables/` in alphabetical order. They also
cross-check the file list against the TOT and warn you about any discrepancies.

```{r run-all-figures, message=FALSE, warning=FALSE}
run_all_figures()
```

```{r run-all-tables, message=FALSE, warning=FALSE}
run_all_tables()
```

---

## 9. Generating the report

### Create the report scaffold

`create_report()` reads the TOT and generates a Quarto file at `report/report.qmd`
with a code chunk for every figure and table, in the order they appear in the TOT.

```{r create-report}
create_report(output_type = "html")
```

The generated file looks like this (excerpt):

```{r show-report, echo=FALSE, comment=""}
report_lines <- readLines(fs::path(tmp_proj_dir, "report", "html_report.qmd"))
cat(report_lines[seq_len(min(40L, length(report_lines)))], sep = "\n")
```

### Render the report

Call `run_report()` to render the Quarto file to HTML (or DOCX):

```{r run-report, eval=FALSE}
run_report()
```

After reviewing the HTML report, archive it to `results/reports/` with a
date-stamped filename using `archive_report()`:

```{r archive-report, eval=FALSE}
archive_report()
```

---

## Summary

Here is the full `lbstproj` workflow at a glance:

```{r workflow-summary, eval=FALSE}
library(lbstproj)

# 1 — Scaffold the project
create_project(path = ".", title = "My project", author = "Jane Doe")

# 2 — Fill in data/tot/table_of_tables.xlsx, then import it
import_tot()

# 3 — Generate script stubs for every TOT entry
create_from_tot(dry_run = FALSE)

# 4 — Fill in the scripts, then run them all
run_all_files("data")
run_all_figures()
run_all_tables()

# 5 — Build and render the report
create_report(output_type = "html")
run_report()
archive_report()
```

